---
title: "Decision Trees Assignment"
author: "Abed, Nevin, and Heshan"
date: "2025-04-22"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: true
    toc_depth: 3
    theme: cosmo
    toc_float: true
---


```{r setup, include=FALSE}
if (rstudioapi::isAvailable()) {setwd(dirname(rstudioapi::getActiveDocumentContext()$path))}

options(width = 200)

using <- \(pkg) { # Create our function "using()" 
  if (!rlang::is_installed(pkg)) {  # if a package is not installed, install it
    install.packages(pkg, repo = "https://cloud.r-project.org")}# load the package
  library(pkg, character.only = TRUE)}

using("data.table") # The data manipulation king
using("caret") # The ML swiss-knife - http://topepo.github.io/caret/available-models.html
using("plotly") # Beautiful interactive plots
using("ranger") # the fastest and better random forest implementation
using("xgboost") # Extreme Gradient boosting
using("imbalance") # Oversampling
using("ROSE") # Synthetic generation of new data to rebalance
using("VIM")
using ("ggplot2")
```


```{r data}
data <- fread("data.csv")
```


```{r}
str(data)
#View(data)

```


```{r}
colnames(data)[1:42] <- c(
  "duration", "protocol_type", "service", "flag", "src_bytes", "dst_bytes",
  "land", "wrong_fragt", "urgent", "hot", "num_fail_login", "logged_in",
  "nu_comprom", "root_shell", "su_attempted", "num_root", "nu_file_creat",
  "nu_shells", "nu_access_files", "nu_out_cmd", "is_host_login",
  "is_guest_login", "count", "srv_count", "serror_rate", "srv_serror_rate",
  "rerror_rate", "srv_rerror_rate", "same_srv_rate", "diff_srv_rate",
  "srv_diff_h_rate", "host_count", "host_srv_count", "h_same_sr_rate",
  "h_diff_srv_rate", "h_src_port_rate", "h_srv_d_h_rate", "h_serror_rate",
  "h_sr_serror_rate", "h_rerror_rate", "h_sr_rerror_rate","class")
```


NA checks
```{r}
#View(colSums(is.na(data))) #
sum(complete.cases(data)) # 22469
nrow(data) # 528603
```




Handling duplicates:
```{r}
#sum(duplicated(data)) # 128935
#nrow(data) - sum(duplicated(data))# 399668
data <- data[!duplicated(data), ]
```


Factor all to be factored
```{r}
data$class <- as.factor(data$class)
data$protocol_type <- as.factor(data$protocol_type)
data$service <- as.factor(data$service)
data$flag <- as.factor(data$flag)
```


How many unique values in each factor feature?
```{r}
length(unique(data$flag))
length(unique(data$service))
length(unique(data$protocol_type))
```
> length(unique(data$flag))
[1] 13
> length(unique(data$service))
[1] 68
> length(unique(data$protocol_type))
[1] 5


Having found no usefulness to samples with empty class, we decided remove them.
```{r}
# data[data$class == "", ] |> head(10)
# nrow(data[data$class == "", ]) # 36061
data <- data[data$class != '']
```


#imputing:
#```{r}
#for (j in names(data)) {
#  if(is.numeric(data[[j]])) {
#    data[is.na(get(j)), (j) := median(data[[j]], na.rm = TRUE)]}} # median imputation values is done now 
#nrow(data) # 399668 after removing duplicates
#```


Removing outliers:
Can't remove it right now because the target detection is not binary, but about 26 classes.
The following chunk is disabled for "run above" button
```{r, include=FALSE}
# removing outliers:
#training_without_outliers <- copy(data)
#numeric_cols <- names(data)[sapply(data, is.numeric)]
#for(i in numeric_cols) { 
#  for(j in unique(data$class)) {
#    values <- data[class == j, get(i)]
#    Q1 <- quantile(values, 0.25)
#    Q3 <- quantile(values, 0.75)
#    IQR <- Q3 - Q1
#    lower_bound <- Q1 - 1.5 * IQR
#    upper_bound <- Q3 + 1.5 * IQR
#    training_without_outliers <- training_without_outliers[!(
#      class == j & (get(i) < lower_bound | get(i) > upper_bound))]}}
#
##nrow(training_without_outliers)/nrow(data)
## 0.4350536
```





Handling missing char/factor values:
1. Check for empty cells in each factor column
```{r}
# Count empty cells in each column
empty_cells <- sapply(data, function(x) sum(x == ""))
View(as.data.frame(empty_cells))
```


2. Checking for overlapping values:
```{r}
# Count rows with missing values in each combination of factors
missing_protocol <- data$protocol_type == ""
missing_service <- data$service == ""
missing_flag <- data$flag == ""

# Check total rows with at least one missing value
total_missing_rows <- sum(missing_protocol | missing_service | missing_flag)

# Check overlaps
protocol_service_overlap <- sum(missing_protocol & missing_service)
protocol_flag_overlap <- sum(missing_protocol & missing_flag)
service_flag_overlap <- sum(missing_service & missing_flag)
all_three_missing <- sum(missing_protocol & missing_service & missing_flag)
percent_affected <- (total_missing_rows / nrow(data)) * 100
```
Since removing ~25% of our data would be a substantial loss of information, AI recommends a combined approach:
Analyze the patterns of missingness: Check if the missing values are related to specific classes or patterns in the data. This might give insight into whether the missing data is Missing at Random (MAR) or Missing Not at Random (MNAR).
Mode imputation for rows with a single missing value: For rows where only one of the three variables is missing, use mode imputation.
Consider removal for rows with multiple missing values: For rows where two or all three variables are missing (which appears to be a much smaller subset), consider removal if those patterns don't conform to an identifiable subset of your data.


3. Checking percentages of missing chars in each feature:
```{r}
empty_cells_percentage <- (empty_cells / nrow(data)) * 100
#print(empty_cells_percentage)
#View(as.data.frame(empty_cells_percentage))

empty_cells_df <- data.frame(
  Column = names(empty_cells),
  Empty_Cells = empty_cells,
  Percentage = empty_cells_percentage)
print(empty_cells_df)
```


4. Sorting by percentage of empty cells
```{r, warning=FALSE}
empty_cells_df <- empty_cells_df[order(-empty_cells_df$Percentage), ]
print(empty_cells_df)
```


```{r}
# Create a table of unique values for each feature
unique_values <- sapply(data, function(x) length(unique(x)))
unique_values_df <- data.frame(
  Feature = names(unique_values),
  Unique_Values = unique_values)

# Sort by number of unique values
unique_values_df <- unique_values_df[order(unique_values_df$Unique_Values), ] # Ascending order
# Print the table
print(unique_values_df)

# View in a more readable format
View(unique_values_df)
```


## Addressing Data Leakage and Implementing Proper ML Pipeline

### Step 1: Initial Data Split (Before Any Processing)

```{r, warning=FALSE}
# Set seed for reproducibility
set.seed(42)
# First, shuffle the data before splitting
data <- data[sample(1:nrow(data)), ]
# Second, create a proper initial split BEFORE any preprocessing
# This ensures test data remains completely unseen during training
initial_split_indices <- createDataPartition(data$class, p = 0.7, list = FALSE)
train_initial <- data[initial_split_indices]
test_initial <- data[-initial_split_indices]

cat("\n","Initial training set size:", nrow(train_initial), "\n")
cat("Initial test set size:", nrow(test_initial), "\n")
```
70% was recommended because our data turned out to be not even that big! wtf is then huh? mf 🤨
*The first warning* "Some classes have no records ( )" suggests there are empty/blank values in your class variable that are being treated as a separate class level, and these will be ignored during stratification.
*The second warning* "Some classes have a single record ( target )" indicates there's only one instance of a class level called "target", and this single record will be automatically included in your sample.


### Step 2: Investigating Feature Correlations and Potential Leakage
```{r, include=FALSE}
# Check correlation between numeric features
numeric_cols <- names(train_initial)[sapply(train_initial, is.numeric)]
correlation_matrix <- cor(train_initial[, ..numeric_cols], use = "pairwise.complete.obs")

# Find highly correlated features (above 0.9)
high_correlations <- which(abs(correlation_matrix) > 0.9 & abs(correlation_matrix) < 1, arr.ind = TRUE)
high_correlations_df <- data.frame(
  Feature1 = numeric_cols[high_correlations[, 1]],
  Feature2 = numeric_cols[high_correlations[, 2]],
  Correlation = correlation_matrix[high_correlations])

# Sort by absolute correlation value
high_correlations_df <- high_correlations_df[order(-abs(high_correlations_df$Correlation)), ]
print("Highly correlated features (potential source of leakage):")
print(high_correlations_df)

# Visualize correlation matrix
using("corrplot")
corrplot(correlation_matrix, method = "color", type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.7)

# Check relationship between categorical features and target
# Function to plot categorical feature distribution by class for top classes
plot_cat_distribution <- function(feature, data) {
  # Get top 5 classes
  top_classes <- names(sort(table(data$class), decreasing = TRUE)[1:5])
  
  # Create a subset with only top classes for visualization clarity
  plot_data <- data[data$class %in% top_classes]
  
  # Plot distribution
  ggplot(plot_data, aes_string(x = feature, fill = "class")) +
    geom_bar(position = "fill") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste("Distribution of", feature, "by Class"),
         y = "Proportion")}

# Plot distributions for each categorical feature
plot_cat_distribution("protocol_type", train_initial)
plot_cat_distribution("service", train_initial)
plot_cat_distribution("flag", train_initial)
```
```{r, include=FALSE}


high_correlations_df
"
           Feature1         Feature2 Correlation
1         src_bytes        src_bytes   1.0000000
2       wrong_fragt      wrong_fragt   1.0000000
3    num_fail_login   num_fail_login   1.0000000
7         nu_shells        nu_shells   1.0000000
8   nu_access_files  nu_access_files   1.0000000
24    same_srv_rate    same_srv_rate   1.0000000
33    h_serror_rate    h_serror_rate   1.0000000
38 h_sr_serror_rate h_sr_serror_rate   1.0000000
4         logged_in        logged_in   1.0000000
26       host_count       host_count   1.0000000
34 h_sr_serror_rate    h_serror_rate   0.9876878
37    h_serror_rate h_sr_serror_rate   0.9876878
12  srv_serror_rate      serror_rate   0.9875461
15      serror_rate  srv_serror_rate   0.9875461
17 h_sr_serror_rate  srv_serror_rate   0.9866230
36  srv_serror_rate h_sr_serror_rate   0.9866230
13    h_serror_rate      serror_rate   0.9859460
31      serror_rate    h_serror_rate   0.9859460
16    h_serror_rate  srv_serror_rate   0.9854280
32  srv_serror_rate    h_serror_rate   0.9854280
14 h_sr_serror_rate      serror_rate   0.9847576
35      serror_rate h_sr_serror_rate   0.9847576
18  srv_rerror_rate      rerror_rate   0.9695355
21      rerror_rate  srv_rerror_rate   0.9695355
41 h_sr_rerror_rate    h_rerror_rate   0.9606830
44    h_rerror_rate h_sr_rerror_rate   0.9606830
19    h_rerror_rate      rerror_rate   0.9573980
39      rerror_rate    h_rerror_rate   0.9573980
23 h_sr_rerror_rate  srv_rerror_rate   0.9572933
43  srv_rerror_rate h_sr_rerror_rate   0.9572933
20 h_sr_rerror_rate      rerror_rate   0.9547408
42      rerror_rate h_sr_rerror_rate   0.9547408
22    h_rerror_rate  srv_rerror_rate   0.9539860
40  srv_rerror_rate    h_rerror_rate   0.9539860
27   h_same_sr_rate   host_srv_count   0.9532721
29   host_srv_count   h_same_sr_rate   0.9532721
5          num_root       nu_comprom   0.9504814
6        nu_comprom         num_root   0.9504814
9         srv_count            count   0.9240053
10            count        srv_count   0.9240053
25   h_same_sr_rate    same_srv_rate   0.9112884
28    same_srv_rate   h_same_sr_rate   0.9112884
11  h_src_port_rate        srv_count   0.9108083
30        srv_count  h_src_port_rate   0.9108083"
```


Showing class balance:
```{r}
class_counts <- table(data$class)
class_counts
View(as.data.frame(class_counts))
```
The class imbalance we're seeing (with classes like b'smurf.', b'normal.', and b'neptune.' dominating while others have very few instances) is a separate issue from data leakage.
With severe imbalance, a model might achieve high overall accuracy by favoring the majority classes. 
However, high accuracy may mask poor performance on minority classes, and performance metrics like precision, recall, and F1-score are usually more informative in such cases.



NOTE:
We absolutely can proceed with decision trees and random forests without imputing missing values in categorical features like protocol, service_type, and flag. Both these algorithms have built-in mechanisms to handle missing values:
For decision trees:
They can work with missing values by finding surrogate splits
The algorithm can determine which path a missing value should take based on correlations with other features
For random forests:
Each tree in the ensemble can handle missing values differently
The aggregate prediction across many trees often maintains good accuracy despite missing values
### Step 3: Apply Preprocessing Only on Training Data
```{r}
# 3.1 Handle missing values in categorical features
# First, calculate modes from training data only
train_protocol_mode <- names(sort(table(train_initial$protocol_type[train_initial$protocol_type != ""]), 
                                 decreasing = TRUE))[1]
train_service_mode <- names(sort(table(train_initial$service[train_initial$service != ""]), 
                                decreasing = TRUE))[1]
train_flag_mode <- names(sort(table(train_initial$flag[train_initial$flag != ""]), 
                             decreasing = TRUE))[1]

cat("Mode values from training data:\n")
cat("Protocol type mode:", train_protocol_mode, "\n")
cat("Service mode:", train_service_mode, "\n")
cat("Flag mode:", train_flag_mode, "\n")

# 3.2 Create two versions of the training data
# Version 1: Impute with modes from training data
train_imputed <- copy(train_initial)
train_imputed[protocol_type == "", protocol_type := train_protocol_mode]
train_imputed[service == "", service := train_service_mode]
train_imputed[flag == "", flag := train_flag_mode]

# Version 2: Remove rows with empty values
train_removed <- train_initial[protocol_type != "" & service != "" & flag != "", ]

# 3.3 Apply same transformations to test data
test_imputed <- copy(test_initial)
test_imputed[protocol_type == "", protocol_type := train_protocol_mode]
test_imputed[service == "", service := train_service_mode]
test_imputed[flag == "", flag := train_flag_mode]

test_removed <- test_initial[protocol_type != "" & service != "" & flag != "", ]

# 3.4 For numeric features, calculate median on training data only
for (j in numeric_cols) {
  median_value <- median(train_initial[[j]], na.rm = TRUE)
  
  # Apply to training data
  train_imputed[is.na(get(j)), (j) := median_value]
  train_removed[is.na(get(j)), (j) := median_value]
  train_initial[is.na(get(j)), (j) := median_value]
  
  # Apply to test data
  test_imputed[is.na(get(j)), (j) := median_value]
  test_removed[is.na(get(j)), (j) := median_value]
  #test_initial[is.na(get(j)), (j) := median_value]
}

# 3.5 Report sizes after preprocessing
cat("\nAfter preprocessing:\n")
cat("Training set with imputation:", nrow(train_imputed), "\n")
cat("Training set with removal:", nrow(train_removed), "\n")
cat("Test set with imputation:", nrow(test_imputed), "\n")
cat("Test set with removal:", nrow(test_removed), "\n")
```


### Step 4: Feature Selection to Remove Potential Leakage

```{r, warnings=FALSE}
# 4.1 Alternative to decision trees for feature selection

# First train a basic model to get feature importance
set.seed(42)
initial_model <- ranger(
  class ~ .,
  data = train_imputed,
  importance = 'impurity',
  num.trees = 50,  # Reduced for speed
  mtry = floor(sqrt(ncol(train_imputed) - 1)),
  min.node.size = 5,
  verbose = TRUE
)

# Option 2: Use ranger's built-in feature importance directly
importance_df <- data.frame(
  Feature = names(initial_model$variable.importance),
  Importance = initial_model$variable.importance
)
importance_df <- importance_df[order(-importance_df$Importance), ]
print("Top the least important features based on Random Forest importance:")
print(importance_df)
#View(importance_df)

potentially_leaky_features_rf <- names(head(sort(initial_model$variable.importance, decreasing = T), 5))
cat("\nTop 5 potentially leaky features based on Random Forest importance:\n")
print(potentially_leaky_features_rf)

# Option 3: Use correlation to eliminate redundant features
redundant_features <- c()
correlation_threshold <- 0.95
# Skip self-correlations (same feature correlated with itself)
for (i in 1:nrow(high_correlations_df)) {
  # Only consider different features that are highly correlated
  if (high_correlations_df$Correlation[i] > correlation_threshold & 
      high_correlations_df$Feature1[i] != high_correlations_df$Feature2[i]) {
    # Keep the first feature in each highly correlated pair
    redundant_features <- c(redundant_features, high_correlations_df$Feature2[i])
  }
}
redundant_features <- unique(redundant_features)
cat("\nRedundant features to remove based on correlation (threshold", correlation_threshold, "):", length(redundant_features), "\n")
print(redundant_features)

# Combine both approaches - RF important features and correlation redundant features
potentially_leaky_features <- unique(c(potentially_leaky_features_rf, redundant_features))
cat("\nTotal features to consider removing:", length(potentially_leaky_features), "\n")

# Create reduced datasets without potentially leaky features
train_imputed_reduced <- train_imputed[, !potentially_leaky_features, with = FALSE]
test_imputed_reduced <- test_imputed[, !potentially_leaky_features, with = FALSE]
train_removed_reduced <- train_removed[, !potentially_leaky_features, with = FALSE] 
test_removed_reduced <- test_removed[, !potentially_leaky_features, with = FALSE]

# Report dimensions after feature reduction
cat("\nDimensions after feature reduction:\n")
cat("Original features:", ncol(train_imputed), "\n")
cat("Reduced features:", ncol(train_imputed_reduced), "\n")
cat("Features removed:", ncol(train_imputed) - ncol(train_imputed_reduced), "\n")
```


### Step 5: Model Training with Proper Validation
```{r, warning=FALSE}
# 5.1 Train model WITHOUT potentially leaky features
model_reduced <- ranger(
  class ~ ., 
  data = train_imputed_reduced,
  importance = 'impurity',
  num.trees = 100,
  mtry = floor(sqrt(ncol(train_imputed_reduced) - 1)),
  min.node.size = 5,
  verbose = TRUE)

# 5.2 Evaluate on test set
pred_reduced <- predict(model_reduced, data = test_imputed_reduced)
test_acc_reduced <- sum(pred_reduced$predictions == test_imputed_reduced$class) / nrow(test_imputed_reduced)

# 5.3 Compare with original model (with all features)
model_imputed <- ranger(
  class ~ ., 
  data = train_imputed,
  importance = 'impurity',
  num.trees = 100,
  mtry = floor(sqrt(ncol(train_imputed) - 1)),
  min.node.size = 5,
  verbose = TRUE)

pred_full <- predict(model_imputed, data = test_imputed)
test_acc_full <- sum(pred_full$predictions == test_imputed$class) / nrow(test_imputed)

cat("\nAccuracy comparison:\n")
cat("Model with all features:", test_acc_full, "\n")
cat("Model without leaky features:", test_acc_reduced, "\n")
```


### Step 6: Per-Class Performance Analysis
```{r, warning=FALSE}
# Generate confusion matrix
test_conf_matrix <- table(Predicted = pred_reduced$predictions, Actual = test_imputed_reduced$class)

# Calculate per-class metrics
per_class_metrics <- data.frame(
  Class = colnames(test_conf_matrix),
  Precision = diag(test_conf_matrix) / colSums(test_conf_matrix),
  Recall = diag(test_conf_matrix) / rowSums(test_conf_matrix)
)

# Add F1 score
per_class_metrics$F1 <- 2 * (per_class_metrics$Precision * per_class_metrics$Recall) / 
                         (per_class_metrics$Precision + per_class_metrics$Recall)

# Add support (number of instances)
per_class_metrics$Support <- table(test_imputed_reduced$class)[per_class_metrics$Class]

# Replace NaN values with 0
per_class_metrics[is.na(per_class_metrics)] <- 0

# Sort by support (frequency)
per_class_metrics <- per_class_metrics[order(-per_class_metrics$Support), ]

# Display top 10 classes by frequency
head(per_class_metrics, 10)

# Calculate macro average (unweighted mean of all classes)
macro_avg <- data.frame(
  Class = "macro_avg",
  Precision = mean(per_class_metrics$Precision, na.rm = TRUE),
  Recall = mean(per_class_metrics$Recall, na.rm = TRUE),
  F1 = mean(per_class_metrics$F1, na.rm = TRUE),
  Support = sum(per_class_metrics$Support)
)

# Calculate weighted average (weighted by support)
weighted_avg <- data.frame(
  Class = "weighted_avg",
  Precision = sum(per_class_metrics$Precision * per_class_metrics$Support, na.rm = TRUE) / 
              sum(per_class_metrics$Support, na.rm = TRUE),
  Recall = sum(per_class_metrics$Recall * per_class_metrics$Support, na.rm = TRUE) / 
           sum(per_class_metrics$Support, na.rm = TRUE),
  F1 = sum(per_class_metrics$F1 * per_class_metrics$Support, na.rm = TRUE) / 
        sum(per_class_metrics$Support, na.rm = TRUE),
  Support = sum(per_class_metrics$Support)
)

# Combine all metrics
all_metrics <- rbind(per_class_metrics, macro_avg, weighted_avg)
print(all_metrics)
```
1. Strong Evidence of Feature Redundancy
You've successfully identified and removed 17 potentially problematic features (40% of the total), yet your model's accuracy only dropped from 99.92% to 99.71% (a mere 0.21% decrease). This is compelling evidence that:
Many features were indeed redundant
The connection count metrics (count, srv_count) were likely creating data leakage
The model can achieve excellent performance with a much simpler feature set
2. Class Imbalance Issues
The per-class metrics highlight severe class imbalance:
3 classes dominate: smurf (46,345), neptune (31,497), and normal (28,683)
Several classes have fewer than 5 examples
Some attack types have 0% precision/recall (rootkit, ftp_write, loadmodule)
This explains the huge difference between:
Weighted average (99.7%) - dominated by majority classes
Macro average (60.4%) - equally weighted across all classes
3. Feature Importance Analysis
The top features are all traffic volume metrics:
count and srv_count (connection counting)
src_bytes and dst_bytes (data volume)
same_srv_rate (connection patterns)
This suggests the model relies heavily on simple traffic volume features rather than deeper behavioral patterns.
4. Correlation Patterns
The redundant features fall into clear groups:
Error rate metrics: several variants measuring the same thing
Host vs. connection metrics: similar measurements at different levels
Root access indicators: highly correlated exploitation metrics
5. Recommendations
Focus on rare attack detection: Your model performs poorly on uncommon attack types - consider:
Hierarchical classification (first normal/attack, then attack type)
Oversampling rare classes (SMOTE)
Class-weighted loss functions
Further feature engineering: Create more discriminative features for the rarer attack types
Evaluate on challenging subsets: Test the model only on rare attack types to better assess its real-world utility
The fact that removing suspicious features barely hurt performance confirms you were right to investigate data leakage. The model is now more reliable, generalizable, and better represents real-world performance.






















### Step 7: Learning Curves to Diagnose Overfitting

```{r, warning=FALSE}
# 7.1 Create a function to plot learning curves
plot_learning_curve <- function(data, target, model_fn, train_sizes = seq(0.1, 1.0, by = 0.1)) {
  n <- nrow(data)
  train_scores <- numeric(length(train_sizes))
  test_scores <- numeric(length(train_sizes))
  
  # Create a validation set
  set.seed(123)
  indices <- sample(1:n, size = n * 0.8)
  train_full <- data[indices, ]
  validation <- data[-indices, ]
  
  for (i in seq_along(train_sizes)) {
    # Take a subset of the training data
    size <- floor(train_sizes[i] * nrow(train_full))
    train_subset <- train_full[1:size, ]
    
    # Train model
    model <- model_fn(train_subset)
    
    # Evaluate on training and validation
    train_pred <- predict(model, data = train_subset)
    train_scores[i] <- sum(train_pred$predictions == train_subset[[target]]) / nrow(train_subset)
    
    val_pred <- predict(model, data = validation)
    test_scores[i] <- sum(val_pred$predictions == validation[[target]]) / nrow(validation)
  }
  
  # Create plot data
  plot_data <- data.frame(
    TrainSize = train_sizes * nrow(train_full),
    TrainScore = train_scores,
    ValidationScore = test_scores
  )
  
  # Plot
  ggplot(plot_data, aes(x = TrainSize)) +
    geom_line(aes(y = TrainScore, color = "Training Score")) +
    geom_line(aes(y = ValidationScore, color = "Validation Score")) +
    geom_point(aes(y = TrainScore, color = "Training Score")) +
    geom_point(aes(y = ValidationScore, color = "Validation Score")) +
    scale_color_manual(values = c("Training Score" = "blue", "Validation Score" = "red")) +
    labs(title = "Learning Curve",
         x = "Training Set Size",
         y = "Accuracy",
         color = "Score Type") +
    theme_minimal()
}

# 7.2 Define model function for ranger
ranger_fn <- function(data) {
  ranger(
    class ~ ., 
    data = data,
    num.trees = 50,  # Reduced for speed
    mtry = floor(sqrt(ncol(data) - 1)),
    min.node.size = 5
  )
}

# 7.3 Plot learning curve
plot_learning_curve(train_imputed_reduced, "class", ranger_fn)

# Compare with a smaller feature set
# Select just a few key features for a simpler model
key_features <- c("protocol_type", "service", "flag", "src_bytes", "dst_bytes", "logged_in")
train_simple <- train_imputed[, c(key_features, "class"), with = FALSE]

# Plot learning curve with simple model
plot_learning_curve(train_simple, "class", ranger_fn)
```

### Step 8: Recommended Research Questions

Based on our analysis of potential data leakage and class imbalance, here are three focused research questions:

1. **How do protocol metadata features alone (protocol_type, service, flag) perform in detecting network attacks compared to traffic volume metrics?**
   - This question helps isolate the predictive power of categorical network attributes versus potentially leaky volume metrics.

2. **What are the distinctive patterns of minority attack classes that differentiate them from normal traffic and common attack types?**
   - Focuses on understanding rare attacks that might be overlooked due to class imbalance.

3. **Can temporal patterns in network connection features provide early warning signals for specific attack types?**
   - Investigates whether sequence patterns in the data could improve attack prediction before they fully develop.

These questions directly address the issues we've identified with potential data leakage and class imbalance while aligning with the project's core objective of network intrusion detection.

## Implementations for the recommended approach

```{r, eval=FALSE}
# Additional implementations of the approaches you suggested:

# 1. Dimensionality reduction with PCA
using("stats")
# Only apply PCA to numeric features
pca_features <- prcomp(train_imputed[, ..numeric_cols], scale. = TRUE)
# Determine number of components that explain 95% of variance
variance_explained <- cumsum(pca_features$sdev^2 / sum(pca_features$sdev^2))
n_components <- which(variance_explained >= 0.95)[1]
# Create reduced feature set
pca_train <- predict(pca_features, train_imputed[, ..numeric_cols])[, 1:n_components]
pca_test <- predict(pca_features, test_imputed[, ..numeric_cols])[, 1:n_components]

# 2. Class balancing with SMOTE (for minority classes)
# This would be implemented for severely imbalanced classes

# 3. Different model architectures
# - Try simpler models like decision trees
# - Consider ensemble methods like XGBoost
# - Implement neural networks for complex pattern recognition
```

